# Data transformation

```{r}
library(tidyverse)
library(GGally)
```


## Team attributes

From missing data part we notice that the feature "buildUpPlayDribbling" contains NA values when "buildUpPlayDribblingClass" is little. Firstly we need to separately view the boxplot of "buildUpPlayDribbling" by "buildUpPlayDribblingClass"

```{r}
team = read_csv('data/team_attributes.csv')
team %>%
  ggplot(aes(x = buildUpPlayDribbling, y = factor(buildUpPlayDribblingClass, levels = c("Little", "Normal", "Lots")))) +
  geom_boxplot()
```

So different classes have different scores. Because we need to fill NA value in the Little class, we only draw the histogram of little class and we see that 

```{r}
team %>%
  filter(buildUpPlayDribblingClass == 'Little') %>%
  ggplot(aes(x = buildUpPlayDribbling)) +
  geom_histogram(binwidth = 1)
```

The result of shapiro test shows the p value is 0.0001525 so we would like to fill the NA value by the maximum likelihood estimation of the normal distribution (i.e. mean). 

```{r}
m = team %>% filter(buildUpPlayDribblingClass == 'Little') %>% select(buildUpPlayDribbling) %>% drop_na() 
m = m$buildUpPlayDribbling %>% mean()
team$buildUpPlayDribbling[is.na(team$buildUpPlayDribbling)] = m
```